Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-15 ind_files/01_Introduction.tex
Minor phrasing changes

\section{A very brief outline of the rest of this thesis}
This section provides a bullet-point outline for those who are really short of time. A more detailed outline of the thesis is provided in the next section.

This thesis is structured as follows:

\begin{itemize}
	\item \textbf{Part \ref{part_theory}}
	develops mathematical formalism for describing finite fluctuating populations from first principles.
		\begin{itemize}
		\item Chapter \ref{chap_math_background} provides the necessary mathematical background and provides a toy example studying the size of a population of identical individuals that illustrates the major ideas used. If I am required to shoehorn the structure into the `intro-methods-results-discussion' format, then this chapter can be thought of as providing a mathematical `introduction' and illustration of the `methods' that will be used/generalized in the next chapters.
		\item Chapter \ref{chap_BD} develops a formalism for describing the evolution of finite fluctuating populations that vary in one or more \emph{discrete} characters. This yields equations that generalize the classic Price equation and replicator-mutator equation to finite fluctuating populations. This chapter is biologically original but uses well understood mathematics that has been used in biological contexts before.
		\item Chapter \ref{chap_infD_processes} extends the ideas developed in chapter \ref{chap_BD} to populations that vary in a single one-dimensional \emph{quantitative} character and derives a so-called `stochastic field theory' that describes evolution in such populations. This also results in some mathematical equations that may be of independent interest to physicists and applied mathematicians. This chapter is entirely original, both biologically, and (to the best of my knowledge) in its use of certain mathematical notions that generalize some ideas seen in statistical physics.
		\end{itemize}
	\item \textbf{Part \ref{part_summary}} presents the major results of all the formalism developed up until this point. Those who only want the take home messages and are willing to trust my mathematics can skip to this part.
		\begin{itemize}
			\item Chapter \ref{chap_unification} provides a technical summary of the major results by deriving three important stochastic differential equations. These are equations for type frequencies, the population mean value of an arbitrary type-level quantity, and the population variance of an arbitrary type-level quantity, and respectively generalize the replicator-mutator equation, the Price equation, and \cite{lion_theoretical_2018}'s variance equation to finite, stochastically fluctuating populations. These equations predict a directional evolutionary force called `noise-induced selection' that is only seen in finite, fluctuating populations. Some implications for social evolution and community ecology are discussed.
			\item Section \ref{sec_disc_field_eqns} of chapter \ref{chap_unification} discusses the use of the stochastic field equations formulated in chapter \ref{chap_infD_processes}, connects them to some previous biological models, and briefly discusses the use of and knowledge about such mathematical objects in other areas of applied mathematics such as statistical physics. This section is quite technical due to the nature of the subject.
			\item Chapter \ref{sec_disc} provides a non-technical discussion and summary of all the major results, and discusses biological implications, connections with previous studies, and opportunities for future work. This chapter has no equations (!). Thus, readers who do not want any mathematical equations can skip all other chapters and directly read chapter \ref{sec_disc} if they are interested only in the final results and takeaways\footnote{Though I promise I have tried my best to make sure the journey to this destination is also pleasant :)} 
		\end{itemize}
\end{itemize}

@@

This section provides a more detailed outline of the thesis, for those who want more details than the bullet points above provide.

The rest of this thesis is divided into two parts. Part \ref{part_theory} provides all the gory mathematical details in a (hopefully) accessible pedagogical style. Part \ref{part_summary} then presents some major results and discusses their implications and connections with previous studies.

@@

Part \ref{part_summary} summarizes the major results of our formalism and presents some simple equations that can be argued to be `fundamental theorems' of population biology in the sense of~\cite{queller_fundamental_2017}, which together form a `unifying perspective' in the sense of~\cite{lion_theoretical_2018}. These equations reduce to well-known results such as the Price equation, the replicator-mutator equation from evolutionary game theory, and Fisher's fundamental theorem from population genetics in the infinite population limit. For finite populations, these same equations predict a new evolutionary force, `noise-induced selection', that has still not found its way into the standard formal canon of evolutionary biology and whose significance is only recently being recognized~\citep{constable_demographic_2016,mcleod_social_2019,mazzolini_universality_2022, kuosmanen_turnover_2022}. Implications of noise-induced selection are also discussed in part \ref{part_summary}. Readers who are averse to or do not care for mathematical details can safely skip to Chapter \ref{sec_disc} for the major takeaways of this thesis, though I strongly recommend working through the formalism properly if possible.

>>>>

\section{A very brief outline of the rest of this thesis}
This section provides a bullet-point chapter-wise outline for those who are really short of time. A more detailed outline of the thesis is provided in the next section.

This thesis develops a mathematical formalism for describing finite fluctuating populations from first principles and is structured as follows. The rest of this thesis is divided into two parts. Part \ref{part_theory} provides all the gory mathematical details in a (hopefully) accessible pedagogical style. Part \ref{part_summary} then (re)presents some major results from part \ref{part_theory} and discusses their implications and connections with previous studies.

\begin{itemize}
		\item Chapter \ref{chap_math_background} provides the necessary mathematical background and provides a toy example studying the size of a population of identical individuals that illustrates the major ideas used. If I am required to shoehorn the structure into the `intro-methods-results-discussion' format, then this chapter can be thought of as providing a mathematical `introduction' and illustration of the `methods' that will be used/generalized to get `results' in the following chapters.
		\item Chapter \ref{chap_BD} develops a formalism for describing the evolution of finite fluctuating populations that vary in one or more \emph{discrete} characters. This yields equations that generalize the classic Price equation and replicator-mutator equation to finite fluctuating populations.
		\item Chapter \ref{chap_infD_processes} extends the ideas developed in chapter \ref{chap_BD} to populations that vary in a single one-dimensional \emph{quantitative} character and derives a so-called `stochastic field theory' that describes evolution in such populations. This also results in some mathematical equations that may be of independent interest to physicists and applied mathematicians.
		\item Chapter \ref{chap_unification} provides a technical summary of the major results by deriving three important stochastic differential equations. These are equations for type frequencies, the population mean value of an arbitrary type-level quantity, and the population variance of an arbitrary type-level quantity, and respectively generalize the replicator-mutator equation, the Price equation, and \cite{lion_theoretical_2018}'s variance equation to finite, stochastically fluctuating populations. These equations predict a directional evolutionary force called `noise-induced selection' that is only seen in finite, fluctuating populations. Some implications for social evolution and community ecology are discussed. I also briefly discuss the field equation formalism I develop in chapter \ref{chap_infD_processes}.
		\item Chapter \ref{sec_disc} provides a quick summary of the major results and discusses biological implications, connections with previous studies, and opportunities for future work. This chapter has no equations (!). Thus, readers who do not want any mathematical equations can skip all other chapters and directly read chapter \ref{sec_disc} if they are interested only in the final results and takeaways\footnote{Though I promise I have tried my best to make sure the journey to this destination is also pleasant :)} 
\end{itemize}

@@

This section provides a more detailed outline of the thesis including the technical aspects covered, which sections are original, etc.

@@

Part \ref{part_summary} summarizes the major results of the formalism developed in Part \ref{part_theory} and presents some simple equations that can be argued to be `fundamental theorems' of population biology in the sense of~\cite{queller_fundamental_2017}, which together form a `unifying perspective' in the sense of~\cite{lion_theoretical_2018}. These equations reduce to well-known results such as the Price equation, the replicator-mutator equation from evolutionary game theory, and Fisher's fundamental theorem from population genetics in the infinite population limit. For finite populations, these same equations predict a new evolutionary force, `noise-induced selection', that has still not found its way into the standard formal canon of evolutionary biology and whose significance is only recently being recognized~\citep{constable_demographic_2016,mcleod_social_2019,mazzolini_universality_2022, kuosmanen_turnover_2022}. Implications of noise-induced selection are also discussed in part \ref{part_summary}. Readers who are okay with mathematical equations but do not want any intermediate derivations (or just trust my math) can skip to Chapter \ref{chap_unification} for the major equations and formalisms that emerge as important, though I strongly encourage working through the entire formalism properly if possible. Similarly, readers who are averse to or do not care for equations can safely skip to Chapter \ref{sec_disc} for the major takeaways of this thesis.
